Assessment and implementation of resource road simulation methods for projection of cumulative effects of disturbance on wildlife

### 2023-01-20 analysis/paper/paper.Rmd
minor edits to the text and adding things JH asked for in todos

<!-- With the following code you can access and display values from the yml header above. -->
<!-- Link to data here: https://drive.google.com/file/d/179lw6MPK4l3xGmded9kofzJkB6VlJVb9/view?usp=sharing -->

@@

The impacts of roads on wildlife are complex and varied. Mortality risk and stress on or near roads can be elevated due to vehicle collisions, hunting pressure, predation, noise and other stressors, but roads can also provide food, facilitate movement, and provide refuge [@hill_review_2021;@trombulak_review_2000;@fahrig_effects_2009].    
In Canadian forests, there is ample evidence of linear features attracting predators such as bears (*Ursus americanus*, *Ursus arctos*) and wolves (*Canis lupus*) seeking food or easy movement, while repelling prey such as woodland caribou (*Rangifer tarandus caribou*) and moose (*Alces americanus*) [@dickie_corridors_2020;@polfus_identifying_2011;@mumma_predation_2018;@beauchesne_disentangling_2013;@leblond_avoidance_2013;@laurian_interactions_2012]; there is also evidence of predators avoiding features heavily used by humans [@muhly_human_2011;@whittington_towns_2022; @proctor_effects_2020]. Changes in mortality risks, habitat quality, and connectivity associated with anthropogenic disturbance can in turn alter use and movement patterns [@tucker_moving_2018; @apps_factors_2006; @proctor_effects_2020], demography and population viability [@johnson_science_2020; @decesare_linking_2014; @barrientos_lost_2021;@apps_spatial_2013;@pinard_calving_2012;@fahrig_effects_2009; @lochhead_linking_2022; @apps_spatial_2013; @proctor_effects_2020]. Roads can also facilitate spread of invasive species [@riitters_landscape_2017], and enable other development activities [@johnson_growth-inducing_2020].

Increasingly rapid environmental change, improvements in available cyberinfrastructure, and development of open source data, tools and methods have contributed to growing interest in ecological forecasting models [@bodner_bridging_2021; @dietze_ecological_2017; @mcintire_perfict_2022]. In particular, forest landscape models that can account for harvest, fire, and other disturbances are widely and increasingly used to project the cumulative effects of climate change and forest management [@petter_how_2020; @albrich_simulating_2020; @hof_editorial_2021; @tremblay_harvesting_2018;@snell_importance_2018;@daniel_incorporating_2017;@stewart_climate-informed_2023;@micheletti_assessing_2021;@leblond_there_2022; @bouderbala_long-term_2022; @barros_empowering_2023], building on a rich history of spatial-stochastic models of forest growth and succession following disturbance [@sturtevant_understanding_2021]. Despite the importance of roads for wildlife, these models generally do not include road network projections. For example, in a recent collection of articles focussed on “Using Landscape Simulation Models to Help Balance Conflicting Goals in Changing Forests” [@hof_editorial_2021], only 4 of 14 mentioned roads [@haga_scenario_2020; @leston_quantifying_2020; @mikusinski_strengthening_2021; @norris_forecasting_2021], and only one includes a road projection [@leston_quantifying_2020].

Leston et al. (2020) project forest road network development using Patchworks [@spatial_planning_systems_patchworks_nodate], then add fire. This example highlights that roads are also important for harvest planning, and that road network projection methods are integrated into a widely used optimization-based spatial harvest planning tool. However, Patchworks is proprietary closed-source commercial software, and the road network projection algorithm contained therein is difficult to assess or fully integrate into models that include stochastic disturbances [@sturtevant_understanding_2021; @de_pellegrin_llorente_recognizing_2017]. Published methods intended to support strategic planning [@anderson_projecting_2004;@clark_three-stage_2000;@dean_finding_1997] require inputs that are not publicly available, and employ algorithms that are too complex and detailed for application at regional, provincial or national scales. More promising options for projecting forest road network development at large scales include an iterative least cost path method developed by Hardy [-@hardy_forest_2019], and least cost path and minimum spanning tree algorithms used to project roads and reconstruct development history in British Columbia [@lochhead_demo_2018; @lochhead_linking_2022]. The former method is available as QGIS and Landis plugins, and the latter is embedded in a larger open source modelling project that is specific to British Columbia [@muhly_castor_nodate]. These methods have not been formally assessed, compared, published, or implemented in a way that allows for easy integration into the full variety of modelling frameworks used for forecasting forest change in Canada and elsewhere [e.g. @barros_empowering_2023; @daniel_state-and-transition_2016; @scheller_forest_2004].

In this work, we address the need for adequately accurate forest road network projection tools that are also widely and easily applicable by implementing simple iterative least cost path and minimum spanning tree methods in an open source R package. We compare the ability of these methods to reconstruct observed forest road network development since 1990 in a mountainous region of British Columbia using performance metrics relevant to wildlife. The adequacy of a road network projection method depends on the purpose of the projection, and it is important to select performance metrics that reflect the purpose. Common predictors in models of the impacts of roads on wildlife include distance to roads [], buffered disturbance footprint [@johnson_science_2020;@ibisch_global_2016; @lochhead_linking_2022; @polfus_identifying_2011], and density [@apps_factors_2006; @apps_spatial_2013; @serrouya_time_2017]. We expect projecting these measures to be less challenging than projecting the spatial location of each road segment, and we do not expect good projections of road locations from the algorithms considered here. Our goal is to assess the ability of simple iterative least cost path and minimum spanning tree methods to project footprint, distance and density metrics that are commonly used to predict impacts of roads on wildlife.

<!-- TO DO:
- more thorough review of SMC & grizzly papers
- summarize road metric types used in cited papers more systematically.
-->

<!--
See notes here:
https://docs.google.com/document/d/1UQ_iC69QX0qX86NM6zSSQ036CgslN-uCXCdBY-4nU5M/edit
Effects of roads on wildlife include increased mortality from various types of encounters with humans, alterations in behaviour and interactions among predators and prey, and spread of invasive species (Trombulak and Frissel 2001). In boreal forests and coniferous forests of the Columbia and Rocky mountains, notable impacts of logging roads include facilitation of movement by wolves (Mumma et al. 2018; Dickie et al. 2020), attraction of bears to roadside food resources (Roever et al. 2010; Bastille-Rousseau et al. 2011; Dickie et al. 2020), increased mortality caused by humans near roads and trails (Roever et al. 2010), and avoidance by ungulates (Laurian et al. 2015; Dickie et al. 2020). Forest dwelling boreal woodland and mountain caribou avoid roads leading to a loss of available habitat (Polfus et al. 2011; Dickie et al. 2020), and suffer increased mortality near roads (Bastille-Rousseau et al. 2011; Mumma et al. 2018). At a range scale, survival and recruitment of boreal caribou declines with anthropogenic disturbance (Johnson et al. 2020). Eight herds of Southern Mountain Caribou have been extirpated or likely extirpated, and remaining herds are considered Endangered by COSEWIC  and listed as threatened under the Canadian Species at Risk Act (SARA) (Palm et al. 2019). Most boreal caribou populations with ranges that include intensive forest management activity are not self-sustaining (ECCC 2011; Johnson et al. 2020), and boreal caribou are also listed as threatened under SARA.

@@

-->

<!-- One unpublished QGIS plugin by Clement (2021) is available for projecting forest roads and is also included in the analysis of this paper (GitHub - Klemet/ForestRoadNetworkPluginForQGIS: A plugin for QGIS that is able to construct a forest road network based on a cost matrix, polygons to reach, a current road network, the dijkstra algorithm and 3 network-creation heuristics.). -->

@@

We chose the Revelstoke timber supply area (TSA 27), in British Columbia, Canada as an assessment area because forestry is the dominant resource extraction activity in this region and much of the road network was built to access to forest harvest cutblocks (Figure \@ref(fig:study-area)). This allows projection of the forest road network from historic forest harvest data. Approximately 55% of the area is non-forested land where road building is difficult or impossible (e.g. alpine, lakes, swamp, brush, rock) [@hachey_revelstoke_2010], so the area provides opportunities to observed the responses of road projection algorithms to barriers. Revelstoke is in the interior-wet belt of the province where forests are dominated by Englemann spruce (31%), Hemlock (23%), Cedar (22%), and Douglas-fir (18%) [@ministry_of_forests_revelstoke_2021]. The area is home to wildlife species that are sensitive to roads including grizzly bears and Southern Mountain woodland caribou [@hachey_revelstoke_2010]. Eight herds of Southern Mountain caribou have been extirpated or likely extirpated, and remaining herds are considered Endangered by COSEWIC and listed as threatened under the Canadian Species at Risk Act [@palm_long_2020]. Mountain caribou avoid roads leading to a loss of available habitat [@polfus_identifying_2011; @apps_factors_2006], suffer increased mortality near roads [@apps_spatial_2013], and decline in abundance as the footprint of roads and cutblocks increases [@lochhead_linking_2022]. Roads also alter grizzly habitat use, home range selection, movements, population fragmentation, survival, reproductive rates, population density, trends, and conservation status [@proctor_effects_2020].

@@

To spatially define the cost of building and traveling on a forestry road we developed a simple model informed by slope and major barriers (Table 1). Following the Interior Appraisal Manual of BC, we assume the base cost per kilometer of road is a linear function of slope [@government_of_british_columbia_interior_2021], and that road construction across major water bodies is prohibitively expensive (Table \@ref(tab:cost-tbl)). Existing roads are assigned a cost of 0. The cost raster consists of 1703936 pixels with a resolution of 1 ha.

@@

All data sets were accessed using the `bcdata` R package and filtered to the extent of the Revelstoke Timber Supply Area [@teucher_bcdata_2021].

<!-- TO DO: add links to repositories and explain what data is and is not there, and why -->

@@

Simple methods for projecting low-cost routes from an existing road network to target locations (i.e. harvest cutblocks) include simple least cost paths (LCP), dynamic least cost paths (DLCP) and least cost path minimum spanning trees (MST) (Figure \@ref(fig:proj-meth-fig)). We use Dijkstra's shortest path algorithm [-@dijkstra_note_1959] to determine the least cost path between a target location (e.g. centroid of a cutblock) and the existing road network. Least cost paths to each cutblock is determined independently in simple method (LCP). The dynamic least cost paths (DLCP) method evaluates the path to each cutblock in sequence so that earlier paths are considered in the cost for later paths. Sequence is determined by a heuristic such as the distance to the nearest existing road in these examples. The minimum spanning tree (MST) method calculates the minimum cost network to connecting all cutblocks to the road network using Kruskal's algorithm [-@kruskal_shortest_1956]. See Appendix A for a more detailed explanation of these methods. We focus our analysis on the DLCP and MST approaches which yield more realistic road networks (Figure \@ref(fig:proj-meth-fig)). We compared these methods to an alternative iterative least cost path method developed by Hardy [-@hardy_forest_2019].

<!-- TO DO: cite key R packages used in method implementation -->

@@

To make help make these road network projection methods more easily and widely applicable, we implemented the simple least cost path, dynamic least cost path and minimum spanning tree methods in the open source R package 'roads'.

<!-- TO DO: add links to R package documentation & resources -->

@@

## Performance Metrics
In order to assess the performance of the projections, five characteristics of the projected and observed road networks were calculated and compared: 1) Road presence, measured as whether or not roads are present in each pixel; 2) Road density, measured as density of roads in meters per hectare (m/ha); 3) Distance to nearest road, measured as distance from each raster cell to the nearest road; 4) Road disturbance footprint, measured with a 1 km buffer around roads [@ibisch_global_2016]; and 5) Forestry disturbance footprint, measured with a 500 m buffer around roads and cutblocks. The forestry disturbance footprint is a variant of the buffered anthropogenic disturbance metric that is used as an indicator of boreal caribou critical habitat [@johnson_science_2020], and road density is is often used in studies of anthropogenic impacts on wildlife [@beazley_road_2004; @apps_factors_2006; more]. Forestry disturbance footprint and road density were calculated using the *disturbanceMetrics* and *rasterizeLineDensity* functions in the `caribouMetrics` R package [@hughes_cariboumetrics_2022]. Distance to nearest road was calculated using a fast parallel moving window method implemented in the *getDistFromSource* function of the `roads` package. Only the road and forestry disturbance footprints were calculated for the cutblocks only method since the other characteristics are not applicable without roads. All road characteristics were stored in raster format.

@@

<!-- TO DO: update the details on getDistFromSource function - where is it, or where should it be? -->
<!-- TO DO: Does the cutblocks only scenario include the old roads? If no, it should. And if so, it seems like the other metrics could be calculated in the cutblocks only scenario. It would be interesting to have the "no projection" baseline for all the metrics -->

@@

For each projection method variant the aspatial performance varied across the five road network metrics and also varied depending on whether we consider scores within cutblocks or overall scales (Figure \@ref(fig:aspat-perf-fig)). All road network projection method variants accurately predict the forestry disturbance footprint, and slightly overproject the road disturbance footprint within cutblocks. The high density methods overproject the overall road disturbance footprint more than low density methods, and underproject overall distance to roads, but all method variants give reasonably good projections of overall distance to roads. All methods underproject road density, in part because within 1 ha pixels all the methods generate straight road segments. All the methods also underproject road presence, and the problem is more severe for the low density method variants. In general, performance differences among low and high density method variants are larger than differences among projection algorithms. With the coarser resolution cost raster road projections had mean road presence that was closer to the observed, while road density was further from the observed mean and distance to the nearest road was further from the observed mean inside cutblocks but closer for the landscape overall (Figure A-\@ref(fig:figA1)).

<!-- How is the 'cutblocks only' scenario overestimating the overall road disturbance footprint? That doesn't make sense. -->

@@

Difference maps highlight several potential causes of discrepancy between our projections and the observed road network. In most areas the projections remained within a reasonable vicinity of the observed roads and also realistically remained within valley bottoms without cutting across natural barriers such as mountains (Figure \@ref(fig:proj-map-fig)a). However, the cost raster is not sufficiently details to reflect costs of crossing rivers, streams or other wet areas in valley bottoms so the projections cross valleys more than the observed roads which often run parallel to one another on either side of valleys. There were also instances where cutblocks included in the data set appear to be misidentified and are not accessed by the observed road network (Figure \@ref(fig:proj-map-fig)b). In addition, the projected roads were not able to match the switchback pattern of observed roads in steep terrain (Figure \@ref(fig:proj-map-fig)c). All these discrepancies could be addressed with more accurate harvest and cost information.

@@

# performance table
perf_tbl <- matchData_sum %>% 
  pivot_wider(c(metric, sampleDens), names_from = agreement, 
              values_from = count, values_fill = 0) %>% 
  mutate(sensitivity = `Agree roaded`/(`Agree roaded` + `False negative`)
         precision = `Agree roaded`/(`Agree roaded` + `False positive`),
         F_measure = (2*precision*sensitivity)/(precision+sensitivity)) %>% 
  select(metric, sampleDens, sensitivity, precision, F_measure)

@@

perf_tbl %>% 
  pivot_longer(c(sensitivity, precision, F_measure), names_to = "perf_meas",
               values_to = "value") %>% 
  mutate(perf_meas = factor(perf_meas,
                            labels = c("F-measure", "Precision", "Sensitivity")),
         value = value  +0.0001) %>%   # adding so 0 shows in figure
  ggplot(aes(sampleDens, value, fill = sampleDens))+
  geom_col()+
  scale_fill_manual(values = met.brewer(pal_nm, 8) %>% .[c(8, 1:7)],
                    name = "Sampling\nMethod", guide = "none")+
  facet_grid(metric ~ perf_meas)+
  coord_flip()+
  ylim(0, 1.0001)+
  theme_classic()+
  labs(y = "Performance", x = NULL)

@@

As expected, none of the projection methods can accurately project road presence (i.e. the exact locations of roads) (F-Measure: `r perf_tbl_rng["Road presence"]`), and all of the methods provide accurate projections of the forestry disturbance footprint (F-Measure: `r perf_tbl_rng["Forestry disturbance\nfootprint"]`). All methods also provide reasonably good projections of the road disturbance footprint (F-Measure: `r perf_tbl_rng["Road disturbance\nfootprint"]`), though higher sampling densities yield fewer false negatives and more false positives. All projection methods yielded more accurate results than the 'cutblock only' scenario, indicating that any of these projection methods are better than none (Figure \@ref(fig:spat-perf-fig)). The DLCP projection method had slightly lower performance than the MST projection method with the same sampling strategy. The DLCP method was faster to compute however, with a run time of `r dlcp_time` minutes compared to `r mst_time` minutes for the MST method.

@@

<!-- Discussion pts
- simple algorithms, practical to apply over large areas with minimal available information (unlike operational models)
- as expected, not great at projecting exact road locations. sources of discrepancy include switchback, stream crossings, inaccurate harvest locations. Missing switchbacks and too many stream crossings also reduce our total amount of roads. All of these could be improved with more detailed high resolution cost & harvest and landing data. 
- density of landing pts has a larger effect than choice of road projection algorithm. So maybe don't get too worked up about the differences between one algorithm and another. if more accurate projections are required, focus first on refining cost & harvest landing locations.
- Selected revelstoke because it is topographically complex, so a good challenge. However, This example projected new roads over fairly short distances, new landings were always fairly close to existing roads or other landings so there was not a lot of space for the projection to go wrong. The results may be less accurate if the landings were spread further apart. But also less topographically complex landscapes may be simpler to project.
- For some purposes, projections from relatively crude cost models may be adequate. For this example landscape projections of disturbance footprints (500 m and 1km buffer) and distance to road are pretty good (and better than nothing). Density is harder to project partly because none of these algorithms can project variation in density within 100 m pixels. If projection of impacts on wildlife is a goal, it may be helpful to consider including more easily projectable road predictors in wildlife response models (Bodner,). If interested in predictors that we did not consider, see example script. Can evaluate alternative metrics without rerunning the projections.
-->

<!-- TO CONSIDER: provide example script that would make it relatively easy for someone to assess the performance of a new road metric of their choice. They don't need to run the projections to do this - just summarize the outputs using a method of their choice. -->

>>>>

<!-- With the following code you can access and display values from the yml header above. -->

<!-- Link to data here: https://drive.google.com/file/d/179lw6MPK4l3xGmded9kofzJkB6VlJVb9/view?usp=sharing -->

@@

The impacts of roads on wildlife are complex and varied. Mortality risk and stress on or near roads can be elevated due to vehicle collisions, hunting pressure, predation, noise and other stressors, but roads can also provide food, facilitate movement, and provide refuge [@hill_review_2021; @trombulak_review_2000; @fahrig_effects_2009].\
In Canadian forests, there is ample evidence of linear features attracting predators such as bears (*Ursus americanus*, *Ursus arctos*) and wolves (*Canis lupus*) seeking food or easy movement, while repelling prey such as woodland caribou (*Rangifer tarandus caribou*) and moose (*Alces americanus*) [@dickie_corridors_2020; @polfus_identifying_2011; @mumma_predation_2018; @beauchesne_disentangling_2013; @leblond_avoidance_2013; @laurian_interactions_2012]; there is also evidence of predators avoiding features heavily used by humans [@muhly_human_2011; @whittington_towns_2022; @proctor_effects_2020]. Changes in mortality risks, habitat quality, and connectivity associated with anthropogenic disturbance can in turn alter use and movement patterns [@tucker_moving_2018; @apps_factors_2006; @proctor_effects_2020], demography and population viability [@johnson_science_2020; @decesare_linking_2014; @barrientos_lost_2021; @apps_spatial_2013; @pinard_calving_2012; @fahrig_effects_2009; @lochhead_linking_2022; @apps_spatial_2013; @proctor_effects_2020]. Roads can also facilitate spread of invasive species [@riitters_landscape_2017], and enable other development activities [@johnson_growth-inducing_2020].

Increasingly rapid environmental change, improvements in available cyberinfrastructure, and development of open source data, tools and methods have contributed to growing interest in ecological forecasting models [@bodner_bridging_2021; @dietze_ecological_2017; @mcintire_perfict_2022]. In particular, forest landscape models that can account for harvest, fire, and other disturbances are widely and increasingly used to project the cumulative effects of climate change and forest management [@petter_how_2020; @albrich_simulating_2020; @hof_editorial_2021; @tremblay_harvesting_2018; @snell_importance_2018; @daniel_incorporating_2017; @stewart_climate-informed_2023; @micheletti_assessing_2021; @leblond_there_2022; @bouderbala_long-term_2022; @barros_empowering_2023], building on a rich history of spatial-stochastic models of forest growth and succession following disturbance [@sturtevant_understanding_2021]. Despite the importance of roads for wildlife, these models generally do not include road network projections. For example, in a recent collection of articles focussed on "Using Landscape Simulation Models to Help Balance Conflicting Goals in Changing Forests" [@hof_editorial_2021], only 4 of 14 mentioned roads [@haga_scenario_2020; @leston_quantifying_2020; @mikusinski_strengthening_2021; @norris_forecasting_2021], and only one includes a road projection [@leston_quantifying_2020].

Leston et al. (2020) project forest road network development using Patchworks [@spatial_planning_systems_patchworks_nodate], then add fire. This example highlights that roads are also important for harvest planning, and that road network projection methods are integrated into a widely used optimization-based spatial harvest planning tool. However, Patchworks is proprietary closed-source commercial software, and the road network projection algorithm contained therein is difficult to assess or fully integrate into models that include stochastic disturbances [@sturtevant_understanding_2021; @de_pellegrin_llorente_recognizing_2017]. Published methods intended to support strategic planning [@anderson_projecting_2004; @clark_three-stage_2000; @dean_finding_1997] require inputs that are not publicly available, and employ algorithms that are too complex and detailed for application at regional, provincial or national scales. More promising options for projecting forest road network development at large scales include an iterative least cost path method developed by Hardy [-@hardy_forest_2019], and least cost path and minimum spanning tree algorithms used to project roads and reconstruct development history in British Columbia [@lochhead_demo_2018; @lochhead_linking_2022]. The former method is available as QGIS and Landis plugins, and the latter is embedded in a larger open source modelling project that is specific to British Columbia [@muhly_castor_nodate]. These methods have not been formally assessed, compared, published, or implemented in a way that allows for easy integration into the full variety of modelling frameworks used for forecasting forest change in Canada and elsewhere [e.g. @barros_empowering_2023; @daniel_state-and-transition_2016; @scheller_forest_2004].

In this work, we address the need for adequately accurate forest road network projection tools that are also widely and easily applicable by implementing simple iterative least cost path and minimum spanning tree methods in an open source R package. We compare the ability of these methods to reconstruct observed forest road network development since 1990 in a mountainous region of British Columbia using performance metrics relevant to wildlife. The adequacy of a road network projection method depends on the purpose of the projection, and it is important to select performance metrics that reflect the purpose. Common predictors in models of the impacts of roads on wildlife include distance to roads [], buffered disturbance footprint [@johnson_science_2020; @ibisch_global_2016; @lochhead_linking_2022; @polfus_identifying_2011], and density [@apps_factors_2006; @apps_spatial_2013; @serrouya_time_2017]. We expect projecting these measures to be less challenging than projecting the spatial location of each road segment, and we do not expect good projections of road locations from the algorithms considered here. Our goal is to assess the ability of simple iterative least cost path and minimum spanning tree methods to project footprint, distance and density metrics that are commonly used to predict impacts of roads on wildlife.

```{=html}
<!-- TO DO:
- more thorough review of SMC & grizzly papers
- summarize road metric types used in cited papers more systematically.
-->
```
```{=html}
<!--
See notes here:
https://docs.google.com/document/d/1UQ_iC69QX0qX86NM6zSSQ036CgslN-uCXCdBY-4nU5M/edit
Effects of roads on wildlife include increased mortality from various types of encounters with humans, alterations in behaviour and interactions among predators and prey, and spread of invasive species (Trombulak and Frissel 2001). In boreal forests and coniferous forests of the Columbia and Rocky mountains, notable impacts of logging roads include facilitation of movement by wolves (Mumma et al. 2018; Dickie et al. 2020), attraction of bears to roadside food resources (Roever et al. 2010; Bastille-Rousseau et al. 2011; Dickie et al. 2020), increased mortality caused by humans near roads and trails (Roever et al. 2010), and avoidance by ungulates (Laurian et al. 2015; Dickie et al. 2020). Forest dwelling boreal woodland and mountain caribou avoid roads leading to a loss of available habitat (Polfus et al. 2011; Dickie et al. 2020), and suffer increased mortality near roads (Bastille-Rousseau et al. 2011; Mumma et al. 2018). At a range scale, survival and recruitment of boreal caribou declines with anthropogenic disturbance (Johnson et al. 2020). Eight herds of Southern Mountain Caribou have been extirpated or likely extirpated, and remaining herds are considered Endangered by COSEWIC  and listed as threatened under the Canadian Species at Risk Act (SARA) (Palm et al. 2019). Most boreal caribou populations with ranges that include intensive forest management activity are not self-sustaining (ECCC 2011; Johnson et al. 2020), and boreal caribou are also listed as threatened under SARA.

@@

-->
```
<!-- One unpublished QGIS plugin by Clement (2021) is available for projecting forest roads and is also included in the analysis of this paper (GitHub - Klemet/ForestRoadNetworkPluginForQGIS: A plugin for QGIS that is able to construct a forest road network based on a cost matrix, polygons to reach, a current road network, the dijkstra algorithm and 3 network-creation heuristics.). -->

@@

We chose the Revelstoke timber supply area (TSA 27), in British Columbia, Canada as an assessment area because forestry is the dominant resource extraction activity in this region and much of the road network was built to access forest harvest cutblocks (Figure \@ref(fig:study-area)). This allows projection of the forest road network from historic forest harvest data. Approximately 55% of the area is non-forested land where road building is difficult or impossible (e.g. alpine, lakes, swamp, brush, rock) [@hachey_revelstoke_2010], so the area provides opportunities to observe the responses of road projection algorithms to barriers. Revelstoke is in the interior-wet belt of the province where forests are dominated by Englemann spruce (31%), Hemlock (23%), Cedar (22%), and Douglas-fir (18%) [@ministry_of_forests_revelstoke_2021]. The area is home to wildlife species that are sensitive to roads including grizzly bears and Southern Mountain woodland caribou [@hachey_revelstoke_2010]. Eight herds of Southern Mountain caribou have been extirpated or likely extirpated, and remaining herds are considered Endangered by COSEWIC and listed as threatened under the Canadian Species at Risk Act [@palm_long_2020]. Mountain caribou avoid roads leading to a loss of available habitat [@polfus_identifying_2011; @apps_factors_2006], suffer increased mortality near roads [@apps_spatial_2013], and decline in abundance as the footprint of roads and cutblocks increases [@lochhead_linking_2022]. Roads also alter grizzly habitat use, home range selection, movements, population fragmentation, survival, reproductive rates, population density, trends, and conservation status [@proctor_effects_2020].

@@

To spatially define the cost of building and traveling on a forestry road we developed a simple model informed by slope and major barriers (Table 1). Following the Interior Appraisal Manual of BC, we assume the base cost per kilometer of road is a linear function of slope [@government_of_british_columbia_interior_2021], and that road construction across major water bodies is prohibitively expensive (Table \@ref(tab:cost-tbl)). Existing roads are assigned a cost of 0. The cost raster consists of 1,703,936 pixels with a resolution of 1 ha.

@@

All data sets were accessed using the `bcdata` R package and filtered to the extent of the Revelstoke Timber Supply Area [@teucher_bcdata_2021]. These data sets are continuously updated so we have stored the versions used in our analysis in an Open Science Framework repository (add link). We were not granted permission to share the DRA data set so it must be accessed from <https://catalogue.data.gov.bc.ca/dataset/bb060417-b6e6-4548-b837-f9060d94743e>.

<!-- TO DO: add link to OSF repository once public-->

@@

Simple methods for projecting low-cost routes from an existing road network to target locations (i.e. harvest cutblocks) include simple least cost paths (LCP), dynamic least cost paths (DLCP) and least cost path minimum spanning trees (MST) (Figure \@ref(fig:proj-meth-fig)). We use Dijkstra's shortest path algorithm [-@dijkstra_note_1959] to determine the least cost path between a target location (e.g. centroid of a cutblock) and the existing road network. Least cost paths to each cutblock are determined independently in the simple method (LCP). The dynamic least cost paths (DLCP) method evaluates the path to each cutblock in sequence so that earlier paths are considered in the cost for later paths. Sequence is determined by a heuristic, such as the distance to the nearest existing road in these examples. The minimum spanning tree (MST) method calculates the minimum cost network to connect all cutblocks to the road network using Kruskal's algorithm [-@kruskal_shortest_1956]. See Appendix A for a more detailed explanation of these methods. We focus our analysis on the DLCP and MST approaches which yield more realistic road networks (Figure \@ref(fig:proj-meth-fig)). We compared these methods to an alternative iterative least cost path method developed by Hardy [-@hardy_forest_2019].

@@

To help make these road network projection methods more easily and widely applicable, we implemented the simple least cost path, dynamic least cost path and minimum spanning tree methods in the open source R package `roads` [@endicott2023]. The 'roads' package is dependent on several other R packages that made this work possible, including, among others: `igraph` for calculating least cost paths and minimum spanning trees [@csardi2006a; @igraph:2022], `terra` for manipulating spatial raster data [@hijmans2023], and `sf` for manipulating spatial vector data [@pebesma2018; @pebesma2022]. The package is available on CRAN (The Comprehensive R Archive Network) and a website with tutorials and detailed documentation is available (<https://landscitech.github.io/roads/>). The source code for the package is available on GitHub (<https://github.com/LandSciTech/roads>).

@@

## Performance Metrics

In order to assess the performance of the projections, five characteristics of the projected and observed road networks were calculated and compared: 1) Road presence, measured as whether or not roads are present in each pixel; 2) Road density, measured as density of roads in meters per hectare (m/ha); 3) Distance to nearest road, measured as distance from each raster cell to the nearest road; 4) Road disturbance footprint, measured with a 1 km buffer around roads [@ibisch_global_2016]; and 5) Forestry disturbance footprint, measured with a 500 m buffer around roads and cutblocks. The forestry disturbance footprint is a variant of the buffered anthropogenic disturbance metric that is used as an indicator of boreal caribou critical habitat [@johnson_science_2020], and road density is often used in studies of anthropogenic impacts on wildlife [@beazley_road_2004; @apps_factors_2006; more]. Forestry disturbance footprint and road density were calculated using the *disturbanceMetrics* and *rasterizeLineDensity* functions in the `caribouMetrics` R package [@hughes_cariboumetrics_2022]. Distance to nearest road was calculated using a fast parallel moving window method implemented in the *getDistFromSource* function of the `roads` package. All road characteristics were stored in raster format.

@@

@@

For each projection method variant the aspatial performance varied across the five road network metrics and also varied depending on whether we consider scores within cutblocks, outside cutblocks, or overall (Figure \@ref(fig:aspat-perf-fig)). All road network projection method variants perform better for all metrics than the "cutblocks only" method where cutblocks were the only disturbance considered. All road network projection method variants accurately predict the forestry disturbance footprint, and slightly overproject the road disturbance footprint within cutblocks. The high density methods overproject the outside cutblock road disturbance footprint more than low density methods, and underproject the outside cutblock distance to roads, but all method variants give reasonably good projections of the outside cutblock distance to roads. All methods underproject road density, in part because within 1 ha pixels all the methods generate straight road segments. All the methods also underproject road presence, and the problem is more severe for the low density method variants. In general, performance differences among low and high density method variants are larger than differences among projection algorithms. With the coarser resolution cost raster road projections had mean road presence that was closer to the observed, while road density was further from the observed mean and distance to the nearest road was further from the observed mean inside cutblocks but closer for the landscape overall (Figure A-\@ref(fig:figA1)).

@@

Difference maps highlight several potential causes of discrepancy between our projections and the observed road network. In most areas the projections remained within a reasonable vicinity of the observed roads and also realistically remained within valley bottoms without cutting across natural barriers such as mountains (Figure \@ref(fig:proj-map-fig)a). However, the cost raster is not sufficiently detailed to reflect costs of crossing rivers, streams or other wet areas in valley bottoms so the projections cross valleys more than the observed roads which often run parallel to one another on either side of valleys. There were also instances where cutblocks included in the data set appear to be misidentified and are not accessed by the observed road network (Figure \@ref(fig:proj-map-fig)b). In addition, the projected roads were not able to match the switchback pattern of observed roads in steep terrain (Figure \@ref(fig:proj-map-fig)c). All these discrepancies could be addressed with more accurate harvest and cost information.

<!--# TODO: I don't know that the switchback issue could be addressed with more accurate cost information. Wouldn't you need to account for the direction of the slope and cost according to that in order to induce that behaviour? I guess that would be better cost info but is more complicated than this implies.  -->

@@

# performance table
perf_tbl <- matchData_sum %>% 
  pivot_wider(c(metric, sampleDens), names_from = agreement, 
              values_from = count, values_fill = 0) %>% 
  mutate(sensitivity = `Agree roaded`/(`Agree roaded` + `False negative`),
         precision = `Agree roaded`/(`Agree roaded` + `False positive`),
         F_measure = (2*precision*sensitivity)/(precision+sensitivity)) %>% 
  select(metric, sampleDens, sensitivity, precision, F_measure)

@@

perf_tbl %>% 
  pivot_longer(c(sensitivity, precision, F_measure), names_to = "perf_meas",
               values_to = "value") %>% 
  mutate(perf_meas = factor(perf_meas,
                            labels = c("F-measure", "Precision", "Sensitivity")),
         value = value  +0.01) %>%   # adding so 0 shows in figure
  ggplot(aes(sampleDens, value, fill = sampleDens))+
  geom_col()+
  scale_fill_manual(values = met.brewer(pal_nm, 8) %>% .[c(8, 1:7)],
                    name = "Sampling\nMethod", guide = "none")+
  facet_grid(metric ~ perf_meas)+
  coord_flip()+
  ylim(0, 1.01)+
  theme_classic()+
  labs(y = "Performance", x = NULL)

@@

As expected, none of the projection methods can accurately project road presence (i.e. the exact locations of roads) (F-Measure: `r perf_tbl_rng["Road presence"]`), and all of the methods provide accurate projections of the forestry disturbance footprint (F-Measure: `r perf_tbl_rng["Forestry disturbance\nfootprint"]`). All methods also provide reasonably good projections of the road disturbance footprint (F-Measure: `r perf_tbl_rng["Road disturbance\nfootprint"]`), though higher sampling densities yield fewer false negatives and more false positives (Figure \@ref(fig:spat-perf-fig)). The performance of the "cutblocks only" method could not be assessed for road presence or disturbance footprint since no roads were projected. However, the "cutblocks only" method performed better than any other method for the forestry disturbance footprint. This reflected the fact that in the Revelstoke TSA most new roads were built inside  cutblocks, so the number of false negatives caused by not projecting roads was small. The DLCP projection method had slightly lower performance than the MST projection method with the same sampling strategy. The DLCP method was faster to compute however, with a run time of `r dlcp_time` minutes compared to `r mst_time` minutes for the MST method.

@@

```{=html}
<!-- Discussion pts
- simple algorithms, practical to apply over large areas with minimal available information (unlike operational models)
- as expected, not great at projecting exact road locations. sources of discrepancy include switchback, stream crossings, inaccurate harvest locations. Missing switchbacks and too many stream crossings also reduce our total amount of roads. All of these could be improved with more detailed high resolution cost & harvest and landing data. 
- density of landing pts has a larger effect than choice of road projection algorithm. So maybe don't get too worked up about the differences between one algorithm and another. if more accurate projections are required, focus first on refining cost & harvest landing locations.
- Selected revelstoke because it is topographically complex, so a good challenge. However, This example projected new roads over fairly short distances, new landings were always fairly close to existing roads or other landings so there was not a lot of space for the projection to go wrong. The results may be less accurate if the landings were spread further apart. But also less topographically complex landscapes may be simpler to project.
- For some purposes, projections from relatively crude cost models may be adequate. For this example landscape projections of disturbance footprints (500 m and 1km buffer) and distance to road are pretty good (and better than nothing). Density is harder to project partly because none of these algorithms can project variation in density within 100 m pixels. If projection of impacts on wildlife is a goal, it may be helpful to consider including more easily projectable road predictors in wildlife response models (Bodner,). If interested in predictors that we did not consider, see example script. Can evaluate alternative metrics without rerunning the projections.
-->
```
<!-- TO CONSIDER: provide example script that would make it relatively easy for someone to assess the performance of a new road metric of their choice. They don't need to run the projections to do this - just summarize the outputs using a method of their choice. -->
